% ============================================================================
% REFERENCIAS BIBLIOGRAFICAS
% ============================================================================

\section*{REFER\^ENCIAS BIBLIOGR\'AFICAS}
\addcontentsline{toc}{section}{Refer\^encias Bibliogr\'aficas}

\begin{description}[style=nextline, leftmargin=0pt, labelindent=0pt]

\item[ABNT, 2009a.] ASSOCIA\c{C}\~AO BRASILEIRA DE NORMAS T\'ECNICAS. \textbf{NBR ISO 14040:2009} --- Gest\~ao ambiental --- Avalia\c{c}\~ao do ciclo de vida --- Princ\'ipios e estrutura. Rio de Janeiro, 2009.

\item[ABNT, 2009b.] ASSOCIA\c{C}\~AO BRASILEIRA DE NORMAS T\'ECNICAS. \textbf{NBR ISO 14044:2009} --- Gest\~ao ambiental --- Avalia\c{c}\~ao do ciclo de vida --- Requisitos e orienta\c{c}\~oes. Rio de Janeiro, 2009.

\item[Ecoinvent, 2024.] ECOINVENT CENTRE. \textbf{Ecoinvent database v3}. Swiss Centre for Life Cycle Inventories, 2024. Dispon\'ivel em: \url{https://ecoinvent.org/}

\item[GreenDelta, 2024.] GREENDELTA GmbH. \textbf{OpenLCA} --- Open source life cycle assessment software. Berlin, 2024. Dispon\'ivel em: \url{https://www.openlca.org/}

\item[Giusti, 2025.] GIUSTI, Gabriela. \textbf{Development of characterization factors for health effects of particulate matter in Brazil}. 2025. Tese (Doutorado em Planejamento e Uso de Recursos Renov\'aveis) --- Universidade Federal de S\~ao Carlos, Sorocaba, 2025. Dispon\'ivel em: \url{https://repositorio.ufscar.br/handle/20.500.14289/22123}

\item[Hauschild \& Huijbregts, 2015.] HAUSCHILD, Michael Z.; HUIJBREGTS, Mark A. J. (Ed.). \textbf{Life Cycle Impact Assessment}. Springer, 2015. (LCA Compendium --- The Complete World of Life Cycle Assessment).

\item[Huijbregts et al., 2017.] HUIJBREGTS, Mark A. J. et al. ReCiPe2016: a harmonised life cycle impact assessment method at midpoint and endpoint level. \textbf{International Journal of Life Cycle Assessment}, v. 22, p. 138--147, 2017.

\item[Rosenbaum et al., 2008.] ROSENBAUM, Ralph K. et al. USEtox --- the UNEP-SETAC toxicity model: recommended characterisation factors for human toxicity and freshwater ecotoxicity in life cycle impact assessment. \textbf{International Journal of Life Cycle Assessment}, v. 13, p. 532--546, 2008.

\item[Verones et al., 2020.] VERONES, Francesca et al. \textbf{LC-Impact Version 1.0 --- A spatially differentiated life cycle impact assessment approach}. 2020. Dispon\'ivel em: \url{https://lc-impact.eu/}

\item[ANTUNES et al., 2020.] ANTUNES, A. P. M. et al. Environmental assessment of a permeable pavement system used to harvest stormwater for non-potable water uses in buildings. \textbf{Science of the Total Environment}, v. 721, p. 137631, 2020.

\item[AZEVEDO et al., 2017.] AZEVEDO, G. M. S. et al. Environmental assessment of bioethanol from cattle manure. \textbf{ReCiPe Midpoint and Endpoint}, v. 1.06, 2017. % CONFERIR: referência incompleta/incomum no material enviado

\item[BARE, 2011.] BARE, J. C. TRACI 2.0: the tool for the reduction and assessment of chemical and other environmental impacts 2.0. \textbf{Clean Technologies and Environmental Policy}, v. 13, n. 5, p. 687--696, 2011.

\item[BUENO et al., 2023.] BUENO, C. et al. Environmental life cycle assessment of sargassum seaweed destination scenarios in Salinópolis, Brazil. \textbf{Journal of Cleaner Production}, 2023. % CONFERIR: faltam volume, número e paginação

\item[CAMARGO et al., 2019.] CAMARGO, V. M. et al. Environmental performance assessment of Natura's cosmetics. \textbf{Journal of Cleaner Production}, v. 230, p. 849--860, 2019.

\item[CAPAZ et al., 2020.] CAPAZ, R. S. et al. Environmental impacts of renewable jet fuel production from sugarcane, soybean, and residues in Brazil. \textbf{Journal of Cleaner Production}, v. 270, p. 122509, 2020.

\item[CAVALETT et al., 2013.] CAVALETT, O. et al. Environmental assessment of ethanol production from sugarcane and gasoline in Brazil. \textbf{Journal of Cleaner Production}, v. 47, p. 201--213, 2013.

\item[COSTA et al., 2023.] COSTA, A. P. A. et al. Comparative LCA of hybrid off-grid energy systems in Amazonian indigenous communities. \textbf{Renewable Energy}, v. 201, p. 1544--1555, 2023.

\item[DEKKER et al., 2020.] DEKKER, S. C. et al. Global sensitivity analysis of life cycle impact assessment models: the case of ReCiPe2016. \textbf{Environmental Science \& Technology}, v. 54, n. 14, p. 8992--9001, 2020.

\item[DICK et al., 2021.] DICK, M. et al. Environmental impacts of Brazilian beef cattle production in the Amazon, Cerrado, Pampa, and Pantanal biomes. \textbf{Journal of Cleaner Production}, v. 187, p. 553--561, 2021.

\item[DU et al., 2019.] DU, C. et al. Robust multi-criteria weighting in comparative LCA and S-LCA: a case study of sugarcane production in Brazil. \textbf{Journal of Cleaner Production}, v. 218, p. 708--717, 2019.

\item[EC-JRC, 2010.] EUROPEAN COMMISSION; JOINT RESEARCH CENTRE. \textbf{International Reference Life Cycle Data System (ILCD) Handbook: nomenclature and other conventions}. 1. ed. Luxembourg: Publications Office of the European Union, 2010.

\item[EDELEN et al., 2018.] EDELEN, B. M. et al. Harmonization of LCIA and LCI data. \textbf{Journal of Industrial Ecology}, v. 22, n. 4, p. 638--649, 2018.

\item[EDELEN et al., 2022.] EDELEN, B. M. et al. Improving the interoperability of LCA data: frameworks and tools. \textbf{The International Journal of Life Cycle Assessment}, v. 27, n. 3, p. 421--433, 2022.

\item[FANTKE et al., 2017.] FANTKE, P. et al. Characterizing aggregated exposure to primary particulate matter: recommended intake fractions for indoor and outdoor sources. \textbf{Environmental Science \& Technology}, v. 51, n. 16, p. 9089--9100, 2017. DOI: 10.1021/acs.est.7b02589.

\item[FANTKE et al., 2019.] FANTKE, P. et al. Global effect factors for exposure to fine particulate matter. \textbf{Environmental Science \& Technology}, v. 53, n. 12, p. 6855--6868, 2019. DOI: 10.1021/acs.est.9b01800. % CONFERIR: no material enviado também apareceu paginação divergente

\item[FERREIRA; LEITE, 2015.] FERREIRA, C. A.; LEITE, M. G. P. Avaliação do ciclo de vida da produção de minério de ferro no Brasil. \textbf{Journal of Cleaner Production}, v. 104, p. 241--252, 2015.

\item[FERREIRA et al., 2020.] FERREIRA, D. A. et al. LCA of heavy-duty trucks considering differential axles. \textbf{Journal of Cleaner Production}, v. 276, p. 124115, 2020.

\item[FRISCHKNECHT; JOLLIET, 2016.] FRISCHKNECHT, R.; JOLLIET, O. \textbf{Global guidance for life cycle impact assessment indicators: Volume 1}. [S. l.]: UNEP-SETAC Life Cycle Initiative, 2016.

\item[GIUSTI et al., 2022.] GIUSTI, G. et al. Health effects of particulate matter formation in Life Cycle Impact Assessment: critical review and recommendation of models for Brazil. \textbf{The International Journal of Life Cycle Assessment}, v. 27, p. 868--884, 2022. DOI: 10.1007/s11367-022-02068-3.

\item[GIUSTI et al., 2023.] GIUSTI, G. et al. Human health impacts of particulate matter emitted from different milk production systems in Brazil: a regionalized LCA sensitivity analysis. \textbf{The International Journal of Life Cycle Assessment}, v. 28, p. 1466--1480, 2023.

\item[GIUSTI et al., 2025.] GIUSTI, G. Development of characterization factors for health effects of particulate matter in Brazil. \textbf{Thesis, Graduate Program in Planning and Use of Renewable Resources, UFSCar}, 2025.

\item[GRONLUND et al., 2015.] GRONLUND, C. J. et al. Intake fractions of PM2.5 in the US: spatial variability and implications. \textbf{Environmental Science \& Technology}, v. 49, n. 10, p. 6012--6019, 2015.

\item[HAUSCHILD; HUIJBREGTS, 2015.] HAUSCHILD, Michael Z.; HUIJBREGTS, Mark A. J. (ed.). \textbf{Life Cycle Assessment}. Cham: Springer, 2015. (LCA Compendium --- The Complete World of Life Cycle Assessment).

\item[HOFSTETTER, 1998.] HOFSTETTER, P. \textbf{Perspectives in life cycle impact assessment: a structured approach to combine models of the technosphere, ecosphere and valuesphere}. Boston: Kluwer Academic Publishers, 1998.

\item[HUIJBREGTS et al., 2017.] HUIJBREGTS, M. A. J. et al. ReCiPe2016: a harmonised life cycle impact assessment method at midpoint and endpoint level. \textbf{The International Journal of Life Cycle Assessment}, v. 22, n. 2, p. 138--147, 2017.

\item[HUMBERT et al., 2011.] HUMBERT, S. et al. Intake fraction for particulate matter: recommendations for life cycle impact assessment. \textbf{Environmental Science \& Technology}, v. 45, n. 11, p. 4808--4816, 2011.

\item[INTERNATIONAL ORGANIZATION FOR STANDARDIZATION, 2006a.] INTERNATIONAL ORGANIZATION FOR STANDARDIZATION. \textbf{ISO 14040: environmental management --- life cycle assessment --- principles and framework}. Geneva: ISO, 2006.

\item[INTERNATIONAL ORGANIZATION FOR STANDARDIZATION, 2006b.] INTERNATIONAL ORGANIZATION FOR STANDARDIZATION. \textbf{ISO 14044: environmental management --- life cycle assessment --- requirements and guidelines}. Geneva: ISO, 2006.

\item[KARKOUR et al., 2020.] KARKOUR, S. et al. Comparative environmental and economic assessment of power systems in 20 countries. \textbf{Journal of Cleaner Production}, v. 248, p. 119164, 2020.

\item[KOHLBECK et al., 2023.] KOHLBECK, F. et al. Environmental performance of recycled polypropylene for household appliances. \textbf{Resources, Conservation and Recycling}, v. 180, p. 106203, 2023.

\item[LASSIO; HADDAD, 2016.] LASSIO, G. D.; HADDAD, A. N. Life Cycle Impact Assessment of masonry system as inner walls: a case study in Brazil. \textbf{Construction and Building Materials}, v. 94, p. 362--369, 2016.

\item[MACENO et al., 2023.] MACENO, M. et al. LCA of reusable and disposable face masks in Brazil. \textbf{Journal of Environmental Management}, v. 321, p. 115783, 2023.

\item[MENEZES et al., 2022.] MENEZES, A. M. et al. Life cycle assessment of bioGLP production routes. \textbf{Journal of Cleaner Production}, v. 356, p. 131849, 2022.

\item[MIRANDA et al., 2023.] MIRANDA, P. et al. LCA of MOF-based materials for gas capture. \textbf{Sustainable Materials and Technologies}, v. 37, p. e00338, 2023.

\item[MONTEIRO et al., 2021.] MONTEIRO, L. S. et al. Environmental impact of concrete industry in Brazil. \textbf{Journal of Cleaner Production}, v. 295, p. 126362, 2021.

\item[MUNERON et al., 2021.] MUNERON, L. M. et al. A performance-based approach for coupling cradle-to-use LCA with operational energy simulation for calcium silicate and clay bricks in masonry buildings. \textbf{Energy and Buildings}, v. 252, p. 111489, 2021.

\item[MUTEL et al., 2019.] MUTEL, C. L. et al. Overview and recommendations for regionalized life cycle impact assessment. \textbf{The International Journal of Life Cycle Assessment}, v. 24, n. 4, p. 856--865, 2019. DOI: 10.1007/s11367-018-1539-4.

\item[NOTTER et al., 2015.] NOTTER, D. A. et al. Contribution of criteria air pollutants to toxicity-related environmental impacts: a case study for Switzerland and implications for LCA. \textbf{Science of the Total Environment}, v. 529, p. 158--165, 2015.

\item[OBERSCHELP et al., 2020.] OBERSCHELP, C. et al. Globally regionalized monthly life cycle impact assessment of particulate matter. \textbf{Environmental Science \& Technology}, v. 54, n. 24, p. 16028--16038, 2020. DOI: 10.1021/acs.est.0c05691. % CONFERIR: no material enviado também apareceu n. 17 e paginação divergente

\item[PASSON et al., 2023.] PASSON, J. et al. Environmental impact assessment of demilitarization processes in Brazil. \textbf{Science of the Total Environment}, v. 856, p. 159023, 2023.

\item[POTTING; HAUSCHILD, 2006.] POTTING, J.; HAUSCHILD, M. Z. Spatial differentiation in life cycle impact assessment: a decade of method development to increase the environmental realism of LCIA. \textbf{The International Journal of Life Cycle Assessment}, v. 11, n. 1, p. 11--13, 2006. DOI: 10.1065/lca2006.04.005.

\item[RAICV, 2019.] REDE DE AVALIAÇÃO DE IMPACTO DO CICLO DE VIDA. \textbf{Recomendação de modelos de Avaliação de Impacto do Ciclo de Vida para o Contexto Brasileiro}. 1. ed. Brasília, DF: IBICT, 2019.

\item[RESTREPO et al., 2015.] RESTREPO, E. C. et al. Comparative LCA of charcoal production in southern Brazil. \textbf{Journal of Cleaner Production}, v. 87, p. 100--109, 2015.

\item[ROCHA; PENTEADO, 2021.] ROCHA, J. C.; PENTEADO, C. S. Environmental assessment of reverse logistics for electronic waste in Brazil. \textbf{Resources, Conservation and Recycling}, v. 164, p. 105170, 2021.

\item[RODRIGUES et al., 2016.] RODRIGUES, T. O. et al. \textbf{Guia Qualidata: requisitos de qualidade de conjuntos de dados para o banco nacional de inventários do ciclo de vida}. Brasília, DF: IBICT, 2016.

\item[ROSADO et al., 2017.] ROSADO, L. P. et al. Life cycle assessment of natural and mixed recycled aggregate production in Brazil. \textbf{Journal of Cleaner Production}, v. 151, p. 634--642, 2017. % CONFERIR: no material enviado há registro muito semelhante atribuído a 2019

\item[SADHUKHAN, 2022.] SADHUKHAN, J. Environmental performance of electricity systems in 15 countries. \textbf{Renewable and Sustainable Energy Reviews}, v. 162, p. 112397, 2022.

\item[SANTOS et al., 2017a.] SANTOS, S. A. et al. Mitigation of environmental impacts of beef cattle production in southern Brazil: evaluation using farm-based life cycle assessment. \textbf{Journal of Cleaner Production}, v. 162, p. 178--187, 2017.

\item[SANTOS et al., 2017b.] SANTOS, T. P. et al. Prolonged warming of the Brazil Current precedes deglaciations. \textbf{Earth and Planetary Science Letters}, v. 463, p. 1--12, 2017.

\item[SANYÉ-MENGUAL et al., 2022a.] SANYÉ-MENGUAL, E. et al. Linking inventories and impact assessment models for addressing biodiversity impacts: mapping rules and challenges. \textbf{The International Journal of Life Cycle Assessment}, v. 27, p. 813--833, 2022. DOI: 10.1007/s11367-022-02049-6.

\item[SANYÉ-MENGUAL et al., 2022b.] SANYÉ-MENGUAL, E. et al. Harmonization of life cycle inventory nomenclature. \textbf{The International Journal of Life Cycle Assessment}, v. 27, p. 849--861, 2022.

\item[SICV BRASIL, 2024.] SICV BRASIL. \textbf{Banco Nacional de Inventários do Ciclo de Vida}. 2024. Disponível em: \url{https://sicv.acv.ibict.br/}. Acesso em: 20 ago. 2024.

\item[SILVA et al., 2020.] SILVA, C. A. et al. Regionalized inventory data in LCA of public housing: a comparison between two conventional typologies in southern Brazil. \textbf{Journal of Cleaner Production}, v. 238, p. 117869, 2020.

\item[SOUZA et al., 2015.] SOUZA, D. M. et al. LCA of ceramic vs. concrete tiles and bricks. \textbf{Journal of Cleaner Production}, v. 96, p. 38--45, 2015.

\item[SOUZA et al., 2016.] SOUZA, D. M. et al. Comparative life cycle assessment of ceramic brick, concrete brick and cast-in-place reinforced concrete exterior walls. \textbf{Journal of Cleaner Production}, v. 137, p. 70--82, 2016.

\item[SOUZA et al., 2022.] SOUZA, D. M. et al. Biomass supply chains from sugarcane: LCA comparison. \textbf{Renewable Energy}, v. 190, p. 119--130, 2022.

\item[STAFFORD et al., 2016.] STAFFORD, F. N. et al. Life cycle assessment of the production of Portland cement: a Southern Europe case study. \textbf{Journal of Cleaner Production}, v. 126, p. 159--165, 2016.

\item[STEEN, 1999.] STEEN, B. \textbf{A systematic approach to environmental priority strategies in product development (EPS): version 2000 --- models and data of the default method}. Gothenburg: Chalmers University of Technology, 1999.

\item[TALANG; SIRIVITHAYAPAKORN, 2021.] TALANG, N.; SIRIVITHAYAPAKORN, S. Environmental performance of global MSW disposal scenarios. \textbf{Waste Management}, v. 121, p. 258--268, 2021.

\item[TANG et al., 2018.] TANG, L. et al. Development of human health damage factors for PM2.5 based on a global chemical transport model. \textbf{The International Journal of Life Cycle Assessment}, v. 23, n. 11, p. 2300--2310, 2018.

\item[VALENTE et al., 2024.] VALENTE, T. M. et al. Data harmonization and interoperability in LCA: Brazilian perspective. \textbf{Journal of Cleaner Production}, v. 408, p. 137127, 2024.

\item[VAN ZELM et al., 2008.] VAN ZELM, R. et al. Regionalized life cycle impact assessment of air pollution. \textbf{Atmospheric Environment}, v. 42, p. 134--147, 2008.

\item[VAN ZELM et al., 2016.] VAN ZELM, R. et al. Regionalized life cycle impact assessment of air pollution on the global scale: damage to human health and vegetation. \textbf{Atmospheric Environment}, v. 134, p. 129--137, 2016.

\item[WENGER et al., 2012.] WENGER, J. et al. Indoor intake fractions of fine particulate matter: review of influencing factors. \textbf{Indoor Air}, v. 22, n. 5, p. 392--404, 2012.

\item[WENZEL; HAUSCHILD; ALTING, 1997.] WENZEL, H.; HAUSCHILD, M.; ALTING, L. \textbf{Environmental assessment of products: methodology, tools and case studies in product development}. London: Chapman \& Hall, 1997.

\item[ZAPPE et al., 2020.] ZAPPE, J. et al. Carbon Footprint of agricultural production and processing of tobacco (\textit{Nicotiana tabacum}) in southern Brazil. \textbf{Environmental Technology \& Innovation}, v. 18, p. 100625, 2020.

\end{description}
